Adaptive Sequential Model Selection

### 2015-07-07 paper.tex
Rob edited paper

\newcommand*\mystrut{\vrule width0pt height0pt depth1.5ex\relax}
\newcommand{\underlabel}{\underbracket[1pt][.5pt]{\mystrut \quad\;\; \sub \quad\;\; }}
\newcommand{\JTcomment}[1]{{\color{blue}{(JT: \bf \sc #1) }}}
\newcommand{\WFcomment}[1]{{\color{red}{(WF: \bf \sc #1) }}}
\newcommand{\RTcomment}[1]{{\color{red}{(RT: \bf \sc #1) }}}

@@

More generically, we observe data $Y \in \cY$, with unknown sampling distribution $F$. We then use $Y$ to generate an adaptive sequence of $d$ nested models
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
\WFcomment{should probably end with $M_{d-1}(Y)$. need to fix this   throughout paper.}
By ``model,'' we mean a family of probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration.

@@

\section{Sequential Inference}
\label{sec:sequential}

@@

  Then, $\hk$ controls $\Err$ at $\alpha$
  whenever $p_{[d]}$ satisfy~\eqref{eq:indepCond_random_k0}.
\end{proposition}
\begin{proof}
  For any fixed $k_0$ and $q_1,\ldots, q_{k_0}$, we can construct a
  problem with $\tilde p_k \sim \delta_{q_k}$ for $k\leq k_0$ and 
  $\tilde p_k \sim \text{Unif}[0,1]$ for $k > k_0$. Then, for the
  fixed hypotheses 
  $\widetilde H_k:\; \tilde p_k \sim \text{Unif}[0,1]$, we have
  \[
  f(k_0, q_{[k_0]}) 
  = \E\left[ g\left(\hk(\tilde p_1, \ldots, \tilde p_d),
      k_0\right) \right],
  \]
  which is less than or equal to $\alpha$ by assumption.

@@

\section{Conditions for Independence of p-values}\label{sec:pValsIndep}

We say that the  single-step $p$-value $p_k$ is selectively valid for testing $H_k:\; F \in M_{k-1}(Y)$ against $M_\infty \setminus M_{k-1}$ if, for all $F\in M_\infty$,
\[
\P_F(p_k \leq \alpha \mid M_{k-1},\; F\in M_{k-1}) \leqAS \alpha.
\]
Valid single-step selective $p$-values are not necessarily independent on nulls, but there is a simple sufficient condition under which independence on nulls is guaranteed.

@@

\subsubsection{The Sufficient Filtration}\label{sec:suffFilt}

We will concern ourselves mainly with the {\em sufficient filtration} for the path $M_{[k]}$, defined as
\[
\sF_k^{\text{suff}} = \sF(M_{[k]}, T_k),
\]
We will suppress the superscript and use $\sF_k$ to mean $\sF_k^{\text{suff}}$ for the remainder of the article.

@@

\subsubsection{The Subpath Sufficiency Principle}\label{sec:SSP}

@@

\section{Inference for One Step}
\label{sec:singleStep}

@@

\subsection{Selective $p$-Values in Regression}

In linear regression, there are several types of selective test that we might contemplate performing at step $k$. Let $E_{k-1}$ and $E_k$ denote the active sets for models $M_{k-1}$ and $M_k$ respectively. For simplicity we will assume that only one variable is added at step $k$, so that $E_k = E_{k-1} \cup j_k$. It is relatively straightforward to generalize the developments in this section to the case where multiple variables are added at once.

\subsubsection{Saturated-Model Tests}

Many of the selective tests for linear regression use the saturated-model framework. See for example \citet{lockhart2014significance}, \citet{taylor2013tests}, \citet{taylor2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}.

@@

While the test statistic $\eta_k'Y$ is measurable with respect to $\sF_k$, the reference distribution we compare it against is not. Specifically, $\proj_{\eta_k}^\perp Y$ is not measurable with respect to $\sF_k$. As a result, saturated-model $p$-values constructed as in~\eqref{eq:satModel} are in general {\em not} independent on nulls.

\subsubsection{Selected-Model Tests}
Selected-model inference~\citep{fithian2014optimal} 
represents a more powerful option when testing for inclusion of variable $j_k$. While saturated-model inference always uses the same model $M_{\text{sat}}$, selected-model inference is based on the statistical model that is chosen by our selection procedure. In the sequential case this means that at step $k$, we test the selected null model $M(E_{k-1})$, against either the next selected model $M(E_k)$ or the upper model $M_\infty$.

@@

\subsection{Reducing the number of constraints}
Happily, it turns out that the conditioning polytope can be expressed with only $2p$ constraints in forward stepwise regression, as we show here.
This increases the speed of both accept/reject and hit andRun sampling for our problem.

@@

>>>>

\newcommand*\mystrut{\vrule width0pt height0pt depth1.5ex\relax}
\newcommand{\underlabel}{\underbracket[1pt][.5pt]{\mystrut \quad\;\; \sub \quad\;\; }}
\newcommand{\JTcomment}[1]{{\color{blue}{(JT: \bf \sc #1) }}}
\newcommand{\WFcomment}[1]{{\color{red}{(WF: \bf \sc #1) }}}
\newcommand{\RTcomment}[1]{{\color{green}{(RT: \bf \sc #1) }}}

@@

More generically, we observe data $Y \in \cY$, with unknown sampling distribution $F$. We then use $Y$ to generate an adaptive sequence of $d$ nested models
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
\WFcomment{should probably end with $M_{d-1}(Y)$. need to fix this   throughout paper.}
\RTcomment{Will- be more specific here---   A parametric model  with specified nonzero support?}

By ``model,'' we mean a family of probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration.

@@

\section{Sequential Inference from ordered p-values}
\label{sec:sequential}

@@

  Then, $\hk$ controls $\Err$ at $\alpha$
  whenever $p_{[d]}$ satisfy~\eqref{eq:indepCond_random_k0}.
\end{proposition}
\begin{proof}
  For any fixed $k_0$ and $q_1,\ldots, q_{k_0}$, we can construct a
  problem with $\tilde p_k \sim \delta_{q_k}$ for $k\leq k_0$ and 
  $\tilde p_k \sim \text{Unif}[0,1]$ for $k > k_0$. 
  \RTcomment{Will; what are the qk?  you mean we just choose and fix some values qk? Say so.}Then, for the
  fixed hypotheses 
  $\widetilde H_k:\; \tilde p_k \sim \text{Unif}[0,1]$, we have
  \[
  f(k_0, q_{[k_0]}) 
  = \E\left[ g\left(\hk(\tilde p_1, \ldots, \tilde p_d),
      k_0\right) \right],
  \]
  which is less than or equal to $\alpha$ by assumption.

@@

\section{Construction of Independence of p-values}\label{sec:pValsIndep}

From the previous section, we see that our task is to construct p-values with the {\it independent on nulls} property (\ref{eq:indepCond_random_k0}).
In this section we develop a set of sufficient conditions for such a construction. Before doing so, we give an example to motivate
the general theory.

\medskip
\begin{example}[The selective max-Z test]
Consider the forward stepwise path (\ref{eq:forwardDef_start}---\ref{eq:forwardDef_end}) in the case of linear regression, that is 
$\ell(\theta,Y)=||Y-X\theta||^2$.  Assume that the error variance $\sigma^2$ is known.
 Given an active set of variables $E$ in our current model $M$, the partial regression coefficients for predictors not in $E$ are
\begin{equation}
  z_{j,E}(Y) = \frac{(\proj_E^\perp X_j)'Y}
  {\sigma\|\proj_{E}^\perp X_j\|_2}.
\end{equation}
The max-$z$-statistic for active set $E$  is
\begin{equation}\label{eq:maxT}
  U_E(Y) = \max_{j \notin E} \left|t_{j,E}(Y)\right|,
\end{equation}
and we use $U_{E_{k-1}}(Y)$ as our test statistic at step $k$. We compute a p-value for this test
using the conditional law 
\begin{equation}\label{eq:selModel_marg0}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; X_{E_{k-1}}'Y\right),
\end{equation}
If we reject the hypothesis, we add to $E$ the predictor that maximized $U_E(Y)$ and continue.
This is the procedure that used to obtain the p-values in Table \ref{tab:diab}  \RTcomment{ Rob- make sure you fix table 1 to make the correct computation}

Happily,  it turns that the  p-values from this procedure have exactly the desired {\it independent on nulls} property~\eqref{eq:indepCond_random_k0}.
This happens because of some special characteristics of the construction: 
\begin{description}
\item{(a)}  the inner products $X_E'Y$ are complete sufficient statistics for the regression
parameters in each model $M$,
\item{(b)} knowing only the  model $M$ at step $k$ and its sufficient statistics $X'_EY$, we can compute the entire subpath
of models from the steps 1 to $k-1$. 
\end{description}
 In other words, the inner products $X'_EY$ contains  all of the cumulative information used in the selections from
steps 1 to $k-1$. This in turn results from    the choice  of test statistic $U_E(Y)$  and the law (\ref{eq:selModel_marg0}) used to calibrate it.
We call (b)  the {\it Subpath sufficiency principle} (SSP) in Section \ref{sec:SSP}  below.
In contrast, the {\it saturated model} p-values use a law different from (\ref{eq:selModel_marg0}) to compute p-values, one that uses
information not contained in $X'_EY$. It thus violates the SSP property and hence does not yield p-values that are independent on nulls.
We discuss the regression setting in more detail in Section \ref{sec:selective-reg}.  $\square$
\end{example}

\medskip

We now present a general framework for these ideas, culminating in our theorem \ref{thm:suffCond} at the end of this section below.
We say that the  single-step $p$-value $p_k$ is selectively valid for testing $H_k:\; F \in M_{k-1}(Y)$ against $M_\infty \setminus M_{k-1}$ if, for all $F\in M_\infty$,
\[
\P_F(p_k \leq \alpha \mid M_{k-1},\; F\in M_{k-1}) \leqAS \alpha.
\]
Valid single-step selective $p$-values are not necessarily independent on nulls, but there is a simple sufficient condition under which independence on nulls is guaranteed.

@@

\subsection{The Sufficient Filtration}\label{sec:suffFilt}

We will concern ourselves mainly with the {\em sufficient filtration} for the path $M_{[k]}$, defined as
\[
\sF_k^{\text{suff}} = \sF(M_{[k]}, T_k),
\]
\RTcomment{Will- define $M_{[k-1]}$..  it is  the collection $\{ M_1 M_2 \ldots M_{k-1}\}$?}
We will suppress the superscript and use $\sF_k$ to mean $\sF_k^{\text{suff}}$ for the remainder of the article.

@@

\subsection{The Subpath Sufficiency Principle}\label{sec:SSP}

@@

\section{Inference for one step in the path}
\label{sec:singleStep}

@@

\subsection{Selective $p$-Values in Regression}
\label{sec:selective-reg}
In linear regression, there are several types of selective test that we might contemplate performing at step $k$. Let $E_{k-1}$ and $E_k$ denote the active sets for models $M_{k-1}$ and $M_k$ respectively. For simplicity we will assume that only one variable is added at step $k$, so that $E_k = E_{k-1} \cup j_k$. It is relatively straightforward to generalize the developments in this section to the case where multiple variables are added at once.

\subsubsection{Saturated-Model Tests}
Many of the selective tests for linear regression use the saturated-model framework. See for example \citet{lockhart2014significance}, \citet{taylor2013tests}, \citet{taylor2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}.

@@

While the test statistic $\eta_k'Y$ is measurable with respect to $\sF_k$, the reference distribution we compare it against is not. Specifically, $\proj_{\eta_k}^\perp Y$ is not measurable with respect to $\sF_k$. As a result, saturated-model $p$-values constructed as in~\eqref{eq:satModel} are in general {\em not} independent on nulls.
?Collect
\subsubsection{Selected-Model Tests}
Selected-model inference~\citep{fithian2014optimal} 
represents a more powerful option when testing for inclusion of variable $j_k$. While saturated-model inference always uses the same model $M_{\text{sat}}$, selected-model inference is based on the statistical model that is chosen by our selection procedure. In the sequential case this means that at step $k$, we test the selected null model $M(E_{k-1})$, against either the next selected model $M(E_k)$ or the upper model $M_\infty$.

@@

\subsection{Reducing the number of constraints}
Happily, it turns out that the conditioning polytope can be expressed with only $2p$ constraints in forward stepwise regression, as we show here.
This increases the speed of both accept/reject and hit and Run sampling for our problem.

@@

We plan to distribute  open source software, in both the R and Python languages, , for computing the max-t and other tests discussed in this paper.
